\begin{thesisacknowledgments}

I would like to thank my adviser, Professor Kathleen Fisher, for her
mentorship and guidance. During my undergraduate career, she helped
me build basic research skills that have been invaluable for the
work on this thesis. During the work on this thesis she helped
me develop a plan of attack that made my first real research experience
feel less daunting. She provided the key insight required to
determine when it is safe to scan forward over an arbitrary expression,
my favorite optimization presented in this thesis. Finally,
I would like to thank her for her heroic effort to make my technical
writing less terrible, even when our tools insisted on getting in
the way.

I would also like to thank Professor Nathan Foster for his insight and 
help thinking through the soundness of the optimizations presented
in this thesis. Formalizing my half formed thoughts to convey them
to him helped me catch several logic errors.

The programming work associated with this thesis was all done as
an extension to the Rust programming language's regex crate written
by Andrew Gallant. His work served as a rich source of information,
and made it much easier for me to focus on the key contributions
of this thesis. Without him I would have had to work harder, on
less interesting problems.

The denizens of Halligan Hall patiently let me spew half-coherent
regex facts and ideas at them and provided a safe space for
the controlled release of some light Rust evangelism.
Thanks guys!

Finally, I would like to thank my mother, Gillian Davies, for 
her proofreading help, and my Fianc\`e, Nhu Nguyen, for her
support.

\end{thesisacknowledgments}
